---
title: "Cox Proportional Hazards Model"
author: "Halley Deleeuw and Jasmine Sawh"
date: '`r Sys.Date()`'
format:
  html:
    code-fold: true
    toc: true
    toc-depth: 2
    number-sections: true
course: STA 6257 - Advanced Statistical Modeling
bibliography: references.bib
self-contained: true
execute:
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Introduction

The Cox Proportional Hazards (Cox PH) model is a statistical method used
for survival analysis. It is a semi-parametric estimator that focuses on
the *hazard function*, which represents the instantaneous risk of an
event occurring at a given time, given that the subject has survived up
to that time [@cox1972]. It is widely used in fields such as medicine and
epidemiology, where the outcome of interest is the time until an event
such as death, relapse, or re-infection.

The purpose of the Cox PH model is to describe the relationship between
one or more explanatory variables (covariates) and the hazard, while
allowing for censored observations — subjects for whom the event has not
occurred by the end of the study. The response in a survival model has
two parts: an event indicator (did the event happen?) and a measure of
time (how long until the event or until the subject was last observed)
[@kaplan1958].

In this report we apply the Cox PH model to the North Central Cancer
Treatment Group (NCCTG) lung cancer data set [@loprinzi1994], which
records survival times of patients with advanced lung cancer along with
several clinical measurements. Our goal is to identify which patient
characteristics are associated with the risk of death, to estimate the
size of those effects as hazard ratios, and to check whether the
model's proportional-hazards assumption is reasonable for these data.

## Methods

### The hazard function and the Cox model

Survival analysis models the time $T$ until an event. Two related
functions describe the distribution of $T$. The *survival function*
$S(t) = P(T > t)$ gives the probability of surviving beyond time $t$.
The *hazard function*

$$
h(t) = \lim_{\Delta t \to 0} \frac{P(t \le T < t + \Delta t \mid T \ge t)}{\Delta t}
$$

gives the instantaneous rate of the event at time $t$ for subjects still
at risk. The Cox model [@cox1972] specifies the hazard for a subject with
covariate vector $\mathbf{x} = (x_1, \dots, x_p)$ as

$$
h(t \mid \mathbf{x}) = h_0(t)\,\exp\!\left(\beta_1 x_1 + \beta_2 x_2 + \cdots + \beta_p x_p\right).
\tag{1}
$$

Here $h_0(t)$ is the *baseline hazard* — an unspecified function of time
shared by all subjects — and the exponential term scales that baseline up
or down according to the covariates. Because $h_0(t)$ is left
unspecified, the model is called *semi-parametric*: it makes no
assumption about the shape of the baseline hazard over time, only about
how covariates shift it.

### Hazard ratios

The coefficients are estimated by maximizing Cox's *partial likelihood*,
which depends only on the order in which events occur and therefore does
not require estimating $h_0(t)$ [@cox1972]. The quantity of interest is
the *hazard ratio* (HR), $\exp(\beta_j)$, which compares the hazard for a
one-unit increase in covariate $x_j$, holding the other covariates fixed:

- $\exp(\beta_j) > 1$: higher values of $x_j$ **increase** the hazard
  (worse survival);
- $\exp(\beta_j) < 1$: higher values of $x_j$ **decrease** the hazard
  (better survival);
- $\exp(\beta_j) = 1$: $x_j$ has no effect.

### The proportional-hazards assumption

The name of the model comes from its central assumption: the ratio of
hazards between any two subjects is **constant over time**. Equivalently,
covariate effects do not change as time passes. If this assumption is
violated, the estimated hazard ratios can be misleading [@abeysekera2009].
We assess it using the scaled Schoenfeld residuals [@schoenfeld1982]:
under proportional hazards these residuals should show no trend against
time, and the associated test (implemented as `cox.zph` in the
`survival` package [@survival-package]) should be non-significant.

## Analysis and Results

### Data and visualization

The `lung` data set is included in the `survival` package
[@survival-package]. Each row is a patient; the outcome is time from
study entry (in days) to death or censoring. The covariates we consider
are age, sex, and two measures of performance status — the ECOG score
(`ph.ecog`, higher = more impaired) and the physician-rated Karnofsky
score (`ph.karno`, higher = better) — along with six-month weight loss
(`wt.loss`).

```{r}
#| label: setup
library(survival)
library(ggplot2)
library(knitr)

theme_set(theme_bw(base_size = 12))

# 'lung' ships with the survival package. status is coded 1 = censored,
# 2 = dead; recode to the 0/1 event indicator the models expect, and give
# sex readable labels.
lung$status_event <- lung$status - 1
lung$sex_f <- factor(lung$sex, levels = c(1, 2), labels = c("Male", "Female"))

kable(
  head(lung[, c("time", "status", "age", "sex_f", "ph.ecog", "ph.karno", "wt.loss")]),
  caption = "First six patients in the NCCTG lung cancer data set."
)
```

The data set contains `r nrow(lung)` patients, of whom
`r sum(lung$status_event)` died during follow-up (the remainder were
censored). We begin by estimating the survival function with the
Kaplan–Meier method [@kaplan1958], overall and separately by sex.

```{r}
#| label: km-curve
#| fig-cap: "Kaplan–Meier survival curves by sex. Females (blue) survive longer than males (red)."
# Helper: turn a survfit object into a tidy data frame for ggplot.
tidy_surv <- function(fit) {
  strata <- if (is.null(fit$strata)) {
    rep("All", length(fit$time))
  } else {
    rep(sub(".*=", "", names(fit$strata)), fit$strata)
  }
  # Prepend the (time = 0, surv = 1) origin for each stratum so the step
  # curves start at 100% survival.
  origin <- data.frame(time = 0, surv = 1, strata = unique(strata))
  rbind(origin, data.frame(time = fit$time, surv = fit$surv, strata = strata))
}

km_sex <- survfit(Surv(time, status_event) ~ sex_f, data = lung)

ggplot(tidy_surv(km_sex), aes(time, surv, color = strata)) +
  geom_step(linewidth = 0.9) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  scale_color_manual(values = c(Male = "#c0392b", Female = "#2c7fb8")) +
  labs(x = "Time (days)", y = "Survival probability", color = "Sex",
       title = "Survival by sex — NCCTG lung cancer")
```

The curves separate clearly, with female patients maintaining higher
survival probabilities throughout follow-up. A log-rank test formally
compares the two groups:

```{r}
#| label: logrank
survdiff(Surv(time, status_event) ~ sex_f, data = lung)
```

The log-rank test is highly significant (p ≈ 0.001), confirming that the
difference in survival between males and females is unlikely to be due to
chance. This matches findings in the clinical literature that female lung
cancer patients tend to have better survival than males.

### Statistical modeling

We now fit a Cox proportional hazards model with age, sex, ECOG
performance status, Karnofsky score, and weight loss as covariates.

```{r}
#| label: cox-fit
cox_model <- coxph(
  Surv(time, status_event) ~ age + sex_f + ph.ecog + ph.karno + wt.loss,
  data = lung
)

# Present the estimates as hazard ratios with 95% confidence intervals.
s  <- summary(cox_model)
ci <- s$conf.int
hr_table <- data.frame(
  Term            = rownames(ci),
  `Hazard Ratio`  = round(ci[, "exp(coef)"], 3),
  `Lower 95%`     = round(ci[, "lower .95"], 3),
  `Upper 95%`     = round(ci[, "upper .95"], 3),
  `p-value`       = signif(s$coefficients[, "Pr(>|z|)"], 3),
  check.names = FALSE
)
kable(hr_table, row.names = FALSE,
      caption = "Estimated hazard ratios from the Cox model.")
```

The model as a whole is highly significant (likelihood-ratio test
p ≈ 3 × 10⁻⁶), with a concordance of
`r round(s$concordance[["C"]], 2)`, meaning the model correctly orders
the survival of two randomly chosen patients about
`r round(100 * s$concordance[["C"]])`% of the time. Interpreting the
individual hazard ratios:

- **Sex.** Female patients have a hazard ratio of about
  `r round(ci["sex_fFemale", "exp(coef)"], 2)`, i.e. roughly a
  `r round(100 * (1 - ci["sex_fFemale", "exp(coef)"]))`% lower risk of
  death than males with the same age and performance status
  (p < 0.001). This is the strongest protective effect in the model.
- **ECOG performance status.** Each one-point increase in `ph.ecog`
  (worse functional status) is associated with a hazard ratio of about
  `r round(ci["ph.ecog", "exp(coef)"], 2)` — more than doubling the risk
  of death (p < 0.001).
- **Age, Karnofsky score, and weight loss.** These covariates are not
  statistically significant at the 0.05 level once sex and ECOG status
  are accounted for, though the point estimates are in the expected
  directions.

### Checking the proportional-hazards assumption

Finally we test whether the proportional-hazards assumption holds, using
the scaled Schoenfeld residuals [@schoenfeld1982].

```{r}
#| label: ph-test
ph_test <- cox.zph(cox_model)
ph_test
```

```{r}
#| label: ph-plot
#| fig-cap: "Scaled Schoenfeld residuals for ECOG performance status against time. A roughly flat trend supports the proportional-hazards assumption."
plot(ph_test[3], main = "Schoenfeld residuals: ECOG performance status")
abline(h = 0, col = "red", lty = 2)
```

The **global** test is not significant (p ≈ 0.21), so overall the
proportional-hazards assumption is reasonable for this model. The only
covariate showing a possible violation is the Karnofsky score
(p ≈ 0.03); in practice this could be addressed by stratifying on
Karnofsky score or by adding a time-varying effect, but the global fit is
acceptable for our purposes.

## Conclusion

Using the NCCTG lung cancer data, the Cox proportional hazards model
identified two strong and statistically significant predictors of
survival: **sex** and **ECOG performance status**. Female patients had
roughly half the hazard of death of male patients, and each step of
worsening ECOG status more than doubled the hazard. Age, Karnofsky
score, and recent weight loss were not significant once these two
factors were included. Diagnostic testing with scaled Schoenfeld
residuals indicated that the proportional-hazards assumption held
globally, giving us confidence in the estimated hazard ratios.

These results illustrate the strengths of the Cox model that drew us to
it during our literature review: it handles censored survival data,
produces interpretable hazard ratios without assuming a particular shape
for the baseline hazard, and comes with well-established tools for
checking its central assumption. A natural extension, noted in our
reading on frailty models, would be to account for unobserved
heterogeneity between patients — but for these data the standard Cox
model already provides a clear and well-fitting description of survival.

## References
